---
title: "Getting started with SCOUT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with SCOUT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The `SCOUT` package gives you direct access to the Simulated Cohort of
Universal Tumours from R. 

| Source | What is stored there | Key functions |
|---|---|---|
| **Tables** | Cohort metadata, ground truth tables | `get_metadata()`, `get_ground_truth_cna()`, `get_ground_truth_drivers()`, `get_ground_truth_exposures()`, `get_sampling_information()` |
| **Zenodo** | Sequencing RDS files, normal and tumour sarek results, tumourevo results | `get_sequencing_data()`, `get_normal_data()`, `get_sarek_results()`, `get_tumourevo_results()` |

## Installation

```{r}
devtools::install_github("caravagnalab/SCOUT")
library(SCOUT)
```

---

## Tables

Cohort metadata and ground truth tables are stored as public Google Sheets.

| Function | Description |
|---|---|
| `get_metadata()` | Sample-level annotations (tumour type, clonal class, WGD, sex, …) |
| `get_ground_truth_cna()` | Ground truth copy number segments |
| `get_ground_truth_drivers()` | Ground truth driver events (SNVs, CNAs, WGD) |
| `get_ground_truth_exposures()` | Ground truth mutational signature exposures |
| `get_sampling_information()` | Per-sample clone proportions and sampling time |
| `get_sample_names()` | Sample names for a given SPN |
| `get_tumour_type()` | Tumour type for a given SPN |
| `get_gender()` | Sex chromosome for a given SPN |

All table functions accept optional `spn` and `sample` arguments:

```{r}
get_metadata()
get_ground_truth_cna("SPN01")
get_ground_truth_cna("SPN01", sample = "1.1")
get_ground_truth_drivers("SPN01")
get_ground_truth_exposures("SPN01", type = "SBS")
get_sampling_information("SPN01")

get_sample_names("SPN01")
get_tumour_type("SPN01")
get_gender("SPN01")
```

See the **Tables** article for the full column-level reference.

---

## Zenodo

Data on Zenodo are organised as follows:

| Data type | Content |
|---|---|
| Sequencing ground truth | One record per SPN — `SPN0X_sequencing.tar.gz` |
| Normal sarek outputs | One shared record — `SPN0X_normal.tar.gz` per SPN |
| Sarek + tumourevo (SPN01–06) | One record per SPN per purity (`0.9`, `0.6`, `0.3`) |
| SPN07 sarek | One record per purity |
| SPN07 tumourevo | One record for purity 0.9 + 0.6, one for 0.3 |

For the full list of record IDs see the **Zenodo** article.

Files are downloaded once and cached at `~/.cache/SCOUT/<spn>/`. Override
the cache root with `SCOUT_CACHE_DIR`:

```{r}
Sys.setenv(SCOUT_CACHE_DIR = "/scratch/shared/SCOUT")
```

### Download functions

```{r}
# Tumour sequencing ground truth (all purities and coverages)
get_sequencing_data("SPN04")

# Normal sarek VCF outputs
get_normal_data("SPN04")

# Sarek and tumourevo pipeline results for a given purity
get_sarek_results("SPN04", purity = 0.9)
get_tumourevo_results("SPN04", purity = 0.9)
```

### Getter functions

Once downloaded, dedicated getters resolve file paths without manual
directory navigation:

```{r}
# Ground truth mutations
get_mutations("SPN04", type = "tumour", coverage = 100, purity = 0.9)
get_mutations("SPN04", type = "normal")

# Sarek VCF and CNA files
get_sarek_vcf("SPN04", "SPN04_1.1", 100, 0.9, "mutect2", "tumour")
get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.9, "ascat")

# tumourevo outputs
get_tumourevo_snv("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")
get_tumourevo_cna("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")
get_tumourevo_driver("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza", "mobster", "SPN04_1.1")
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "cnaqc", "SPN04_1.1")
get_tumourevo_signatures("SPN04", 50, 0.6, "mutect2", "sequenza", "BASCULE")
```

See the **Zenodo** article for the full function reference.

---
