---
title: "Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
# the DT table chunk uses eval = TRUE explicitly
```

SCOUT data are distributed across several Zenodo records organised by data
type and SPN. Files are downloaded once and cached locally — repeat calls
detect the cache and skip the download automatically.

## Cache location

By default files are stored at `~/.cache/SCOUT/<spn>/`. Override with:

```{r}
Sys.setenv(SCOUT_CACHE_DIR = "/scratch/shared/SCOUT")
```

---

## Zenodo record structure

```{r zenodo-table, echo = FALSE, eval = TRUE}
records <- data.frame(
  SPN = c(
    rep("all", 1),
    rep("SPN01", 3), rep("SPN02", 3), rep("SPN03", 3),
    rep("SPN04", 3), rep("SPN05", 3), rep("SPN06", 3),
    rep("SPN07", 4)
  ),
  Purity = c(
    "-",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3",
    "0.9", "0.6", "0.3", "0.9 + 0.6"
  ),
  Content = c(
    "Sequencing + normal (all SPNs)",
    rep("sarek + tumourevo", 3),
    rep("sarek + tumourevo", 3),
    rep("sarek + tumourevo", 3),
    rep("sarek + tumourevo", 3),
    rep("sarek + tumourevo", 3),
    rep("sarek + tumourevo", 3),
    "sarek only", "sarek only", "sarek + tumourevo", "tumourevo only"
  ),
  Record_ID = c(
    "18839402",
    "20375904", "20369470", "20201559",
    "20375976", "20362893", "20201684",
    "20376007", "20369483", "20201729",
    "20376032", "20369494", "20201769",
    "20376054", "20369498", "20201784",
    "20376060", "20369500", "20201816",
    "20386282", "20374070", "20201843", "20386301"
  ),
  Archives = c(
    "SPN01_sequencing.tar.gz, SPN01_normal_sequencing.tar.gz, ...",
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    rep("sarek.tar.gz, tumourevo.tar.gz", 3),
    "sarek.tar.gz", "sarek.tar.gz", "sarek.tar.gz, tumourevo.tar.gz",
    "tumourevo_0.9.tar.gz, tumourevo_0.6.tar.gz"
  ),
  stringsAsFactors = FALSE
)

DT::datatable(
  records,
  colnames = c("SPN", "Purity", "Content", "Zenodo record ID", "Archives"),
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 25, dom = "ftip")
)
```

---

## Archive contents

### `SPN0X_sequencing.tar.gz`

Contains tumour mutation RDS files for all purity and coverage combinations.

```
sequencing/
└── tumour/
    ├── purity_0.9/data/mutations/seq_results_muts_merged_coverage_50x.rds
    │                             seq_results_muts_merged_coverage_100x.rds
    │                             seq_results_muts_merged_coverage_150x.rds
    │                             seq_results_muts_merged_coverage_200x.rds
    ├── purity_0.6/data/mutations/  (same files)
    └── purity_0.3/data/mutations/  (same files)
```

### `SPN0X_normal_sequencing.tar.gz`

Contains the matched normal mutation RDS file (fixed at 30x, purity 1).

```
sequencing/
└── normal/
    └── purity_1/data/mutations/seq_results_muts_merged_coverage_30x.rds
```

### `sarek.tar.gz`

Contains Sarek variant calling results for all callers and samples.

```
sarek/
└── <coverage>x_<purity>p/
    └── variant_calling/
        ├── mutect2/<spn>/                         (SPN-level)
        ├── strelka/<sample>_vs_normal_sample/
        ├── freebayes/<sample>_vs_normal_sample/
        │           normal_sample/
        ├── haplotypecaller/normal_sample/
        ├── ascat/<sample>_vs_normal_sample/
        ├── battenberg/<sample>_vs_normal_sample/
        ├── sequenza/<sample>_vs_normal_sample/
        └── cnvkit/<sample>_vs_normal_sample/
```

### `tumourevo.tar.gz`

Contains tumourevo pipeline results for all tools and samples.

```
tumourevo/
└── <coverage>x_<purity>p_<vcf_caller>_<cna_caller>/
    ├── formatter/
    │   ├── vcf2cnaqc/SCOUT/<spn>/<sample>/        (SNV RDS per sample)
    │   ├── cna2cnaqc/SCOUT/<spn>/<sample>/        (CNA RDS per sample)
    │   └── cnaqc2tsv/SCOUT/<spn>/                 (joint table TSV)
    ├── variant_annotation/vep/SCOUT/<spn>/<sample>/
    ├── driver_annotation/annotate_driver/SCOUT/<spn>/<sample>/
    ├── qc/
    │   ├── cnaqc/SCOUT/<spn>/<sample>/
    │   ├── join_cnaqc/SCOUT/<spn>/
    │   └── tinc/SCOUT/<spn>/<sample>/
    ├── subclonal_deconvolution/
    │   ├── mobster/SCOUT/<spn>/<sample>/
    │   ├── viber/SCOUT/<spn>/
    │   ├── pyclonevi/SCOUT/<spn>/
    │   └── ctree/SCOUT/<spn>/  (and per sample)
    └── signature_deconvolution/
        ├── BASCULE/SCOUT/
        ├── sigprofiler/SCOUT/results/<context>/
        └── sparsesignatures/SCOUT/
```

---

## Downloading data

### `get_sequencing_data()`

Downloads `SPN0X_sequencing.tar.gz` from the SPN's sequencing record.
Contains tumour mutation RDS files for all purity and coverage combinations.

```{r}
get_sequencing_data("SPN04")
```

### `get_normal_data()`

Downloads `SPN0X_normal.tar.gz` from the shared normal record.
Contains haplotypecaller and freebayes VCF files for the normal sample.

```{r}
get_normal_data("SPN04")
```

### `get_sarek_results()`

Downloads `sarek.tar.gz` for a given SPN and purity.

```{r}
get_sarek_results("SPN01", purity = 0.9)
get_sarek_results("SPN01", purity = 0.6)
get_sarek_results("SPN01", purity = 0.3)
```

### `get_tumourevo_results()`

Downloads `tumourevo.tar.gz` for a given SPN and purity.

```{r}
get_tumourevo_results("SPN01", purity = 0.9)
```

### `list_zenodo_files()`

Inspect what is available in a record before downloading.

```{r}
list_zenodo_files("1234567")
#> # A tibble: 2 × 3
#>   filename                size download_url
#>   <chr>                  <int> <chr>
#> 1 sarek.tar.gz             ...  https://zenodo.org/...
#> 2 tumourevo.tar.gz         ...  https://zenodo.org/...
```

---

## Ground truth getters

After `get_sequencing_data()` or `get_normal_data()`, use these functions to
access specific files without navigating the directory structure.

### `get_mutations()`

Returns the path to the mutations RDS file for a given type, coverage and
purity.

```{r}
# Tumour (requires coverage and purity)
path <- get_mutations("SPN04", type = "tumour", coverage = 100, purity = 0.9)
readRDS(path)

# Normal (fixed at 30x, purity 1)
path <- get_mutations("SPN04", type = "normal")
```

---

## Sarek getters

After `get_sarek_results()` or `get_normal_data()`, these functions return
named lists of file paths for a given sample, coverage, purity and caller.

### `get_sarek_vcf()`

Returns a named list of VCF and index file paths. Supported callers:
`"mutect2"`, `"strelka"`, `"freebayes"`, `"haplotypecaller"`.
Sample naming convention: `"SPN04_1.1"`.

```{r}
# mutect2 — SPN-level result, no sample needed
vcf <- get_sarek_vcf("SPN04", NULL, 100, 0.3, "mutect2")
vcf$vcf
vcf$tbi

# strelka — one somatic VCF
vcf <- get_sarek_vcf("SPN04", "SPN04_1.1", 100, 0.3, "strelka")
vcf$vcf

# freebayes
vcf <- get_sarek_vcf("SPN04", "SPN04_1.1", 100, 0.3, "freebayes")
vcf$vcf

# haplotypecaller — normal sample (sample = NULL)
vcf <- get_sarek_vcf("SPN04", NULL, 100, 0.3, "haplotypecaller")
vcf$vcf
```

### `get_sarek_cna()`

Returns a named list of CNA file paths. Supported callers: `"ascat"`,
`"battenberg"`, `"sequenza"`, `"cnvkit"`.

```{r}
# ASCAT
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "ascat")
cna$segments
cna$purityploidy
cna$cnvs

# Battenberg
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "battenberg")
cna$subclones
cna$rho_and_psi

# Sequenza
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "sequenza")
cna$segments
cna$confints_CP

# CNVkit
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "cnvkit")
cna$cns
```

---

## tumourevo getters

After `get_tumourevo_results()`, these functions return named lists of file
paths. All require `spn`, `coverage`, `purity`, `vcf_caller` (`"mutect2"` or
`"strelka"`), and `cna_caller` (`"ascat"` or `"sequenza"`). Sample naming
convention: `"SPN04_1.1"`.

### Formatter

```{r}
# Formatted SNV RDS (vcf2cnaqc)
get_tumourevo_snv("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# Formatted CNA RDS (cna2cnaqc)
get_tumourevo_cna("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# Joint CNAqc table (cnaqc2tsv) — one file per combination
get_tumourevo_joint_table("SPN04", 50, 0.6, "mutect2", "sequenza")
```

### Variant annotation

```{r}
# VEP annotated VCF
get_tumourevo_vep("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")
```

### Driver annotation

```{r}
# One sample
get_tumourevo_driver("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# All samples (sample = NULL)
get_tumourevo_driver("SPN04", 50, 0.6, "mutect2", "sequenza")
```

### Subclonal deconvolution

Supported tools: `"mobster"`, `"pyclonevi"`, `"ctree"`, `"viber"`.

```{r}
# mobster — sample required
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza",
                         "mobster", "SPN04_1.1")

# viber — SPN-level result, no sample needed
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza", "viber")

# ctree — SPN-level trees (sample = NULL) or per-sample
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza", "ctree")
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza",
                         "ctree", "SPN04_1.1")
```

### QC

Supported tools: `"cnaqc"`, `"join_cnaqc"`, `"tinc"`.

```{r}
# cnaqc / tinc — sample required
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "cnaqc", "SPN04_1.1")
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "tinc",  "SPN04_1.1")

# join_cnaqc — SPN-level, no sample needed
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "join_cnaqc")
```

### Signature deconvolution

Supported tools: `"sigprofiler"`, `"sparsesignatures"`, `"BASCULE"`.
`sigprofiler` requires a `context` argument (e.g. `"SBS96"`, `"ID83"`,
`"DBS78"`).

```{r}
# BASCULE
sigs <- get_tumourevo_signatures("SPN04", 50, 0.6, "mutect2", "sequenza",
                                  "BASCULE")
sigs$refined_fit
sigs$base_fit

# SigProfiler
sigs <- get_tumourevo_signatures("SPN04", 50, 0.6, "mutect2", "sequenza",
                                  "sigprofiler", context = "SBS96")
sigs$context_matrix
sigs$COSMIC_exposure
sigs$denovo_signatures
```
