---
title: "Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

SCOUT data are distributed across several Zenodo records organised by data
type and SPN. Files are downloaded once and cached locally — repeat calls
detect the cache and skip the download automatically.

## Cache location

By default files are stored at `~/.cache/SCOUT/<spn>/`. Override with:

```{r}
Sys.setenv(SCOUT_CACHE_DIR = "/scratch/shared/SCOUT")
```

---

## Zenodo record structure

| Data type | Zenodo record ID | Archive name |
|---|---|---|
| Sequencing + normal (one shared record, all SPNs) | | `SPN01_sequencing.tar.gz`, `SPN01_normal_sequencing.tar.gz`, … `SPN07_sequencing.tar.gz`, `SPN07_normal_sequencing.tar.gz` |
| SPN01 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN01 sarek + tumourevo purity 0.6 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN01 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN02 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN02 sarek + tumourevo purity 0.6 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN02 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN03 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN03 sarek + tumourevo purity 0.6 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN03 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN04 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN04 sarek + tumourevo purity 0.6 | 505306 | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN04 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN05 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN05 sarek + tumourevo purity 0.6 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN05 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN06 sarek + tumourevo purity 0.9 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN06 sarek + tumourevo purity 0.6 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN06 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN07 sarek purity 0.9 | | `sarek.tar.gz` |
| SPN07 sarek purity 0.6 | | `sarek.tar.gz` |
| SPN07 sarek + tumourevo purity 0.3 | | `sarek.tar.gz`, `tumourevo.tar.gz` |
| SPN07 tumourevo purity 0.9 + 0.6 | | `tumourevo_0.9.tar.gz`, `tumourevo_0.6.tar.gz` |

---

## Downloading data

### `get_sequencing_data()`

Downloads `SPN0X_sequencing.tar.gz` from the SPN's sequencing record.
Contains tumour mutation RDS files for all purity and coverage combinations.

```{r}
get_sequencing_data("SPN04")
```

### `get_normal_data()`

Downloads `SPN0X_normal.tar.gz` from the shared normal record.
Contains haplotypecaller and freebayes VCF files for the normal sample.

```{r}
get_normal_data("SPN04")
```

### `get_sarek_results()`

Downloads `sarek.tar.gz` for a given SPN and purity.

```{r}
get_sarek_results("SPN01", purity = 0.9)
get_sarek_results("SPN01", purity = 0.6)
get_sarek_results("SPN01", purity = 0.3)
```

### `get_tumourevo_results()`

Downloads `tumourevo.tar.gz` for a given SPN and purity.

```{r}
get_tumourevo_results("SPN01", purity = 0.9)
```

### `list_zenodo_files()`

Inspect what is available in a record before downloading.

```{r}
list_zenodo_files("1234567")
#> # A tibble: 2 × 3
#>   filename                size download_url
#>   <chr>                  <int> <chr>
#> 1 sarek.tar.gz             ...  https://zenodo.org/...
#> 2 tumourevo.tar.gz         ...  https://zenodo.org/...
```

---

## Ground truth getters

After `get_sequencing_data()` or `get_normal_data()`, use these functions to
access specific files without navigating the directory structure.

### `get_mutations()`

Returns the path to the mutations RDS file for a given type, coverage and
purity.

```{r}
# Tumour (requires coverage and purity)
path <- get_mutations("SPN04", type = "tumour", coverage = 100, purity = 0.9)
readRDS(path)

# Normal (fixed at 30x, purity 1)
path <- get_mutations("SPN04", type = "normal")
```

---

## Sarek getters

After `get_sarek_results()` or `get_normal_data()`, these functions return
named lists of file paths for a given sample, coverage, purity and caller.

### `get_sarek_vcf()`

Returns a named list of VCF and index file paths. Supported callers:
`"mutect2"`, `"strelka"`, `"freebayes"`, `"haplotypecaller"`.
Sample naming convention: `"SPN04_1.1"`.

```{r}
# mutect2 — SPN-level result, no sample needed
vcf <- get_sarek_vcf("SPN04", NULL, 100, 0.3, "mutect2")
vcf$vcf
vcf$tbi

# strelka — one somatic VCF
vcf <- get_sarek_vcf("SPN04", "SPN04_1.1", 100, 0.3, "strelka")
vcf$vcf

# freebayes
vcf <- get_sarek_vcf("SPN04", "SPN04_1.1", 100, 0.3, "freebayes")
vcf$vcf

# haplotypecaller — normal sample (sample = NULL)
vcf <- get_sarek_vcf("SPN04", NULL, 100, 0.3, "haplotypecaller")
vcf$vcf
```

### `get_sarek_cna()`

Returns a named list of CNA file paths. Supported callers: `"ascat"`,
`"battenberg"`, `"sequenza"`, `"cnvkit"`.

```{r}
# ASCAT
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "ascat")
cna$segments
cna$purityploidy
cna$cnvs

# Battenberg
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "battenberg")
cna$subclones
cna$rho_and_psi

# Sequenza
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "sequenza")
cna$segments
cna$confints_CP

# CNVkit
cna <- get_sarek_cna("SPN04", "SPN04_1.1", 100, 0.3, "cnvkit")
cna$cns
```

---

## tumourevo getters

After `get_tumourevo_results()`, these functions return named lists of file
paths. All require `spn`, `coverage`, `purity`, `vcf_caller` (`"mutect2"` or
`"strelka"`), and `cna_caller` (`"ascat"` or `"sequenza"`). Sample naming
convention: `"SPN04_1.1"`.

### Formatter

```{r}
# Formatted SNV RDS (vcf2cnaqc)
get_tumourevo_snv("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# Formatted CNA RDS (cna2cnaqc)
get_tumourevo_cna("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# Joint CNAqc table (cnaqc2tsv) — one file per combination
get_tumourevo_joint_table("SPN04", 50, 0.6, "mutect2", "sequenza")
```

### Variant annotation

```{r}
# VEP annotated VCF
get_tumourevo_vep("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")
```

### Driver annotation

```{r}
# One sample
get_tumourevo_driver("SPN04", 50, 0.6, "mutect2", "sequenza", "SPN04_1.1")

# All samples (sample = NULL)
get_tumourevo_driver("SPN04", 50, 0.6, "mutect2", "sequenza")
```

### Subclonal deconvolution

Supported tools: `"mobster"`, `"pyclonevi"`, `"ctree"`, `"viber"`.

```{r}
# mobster — sample required
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza",
                         "mobster", "SPN04_1.1")

# viber — SPN-level result, no sample needed
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza", "viber")

# ctree — SPN-level trees (sample = NULL) or per-sample
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza", "ctree")
get_tumourevo_subclonal("SPN04", 50, 0.6, "mutect2", "sequenza",
                         "ctree", "SPN04_1.1")
```

### QC

Supported tools: `"cnaqc"`, `"join_cnaqc"`, `"tinc"`.

```{r}
# cnaqc / tinc — sample required
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "cnaqc", "SPN04_1.1")
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "tinc",  "SPN04_1.1")

# join_cnaqc — SPN-level, no sample needed
get_tumourevo_qc("SPN04", 50, 0.6, "mutect2", "sequenza", "join_cnaqc")
```

### Signature deconvolution

Supported tools: `"sigprofiler"`, `"sparsesignatures"`, `"BASCULE"`.
`sigprofiler` requires a `context` argument (e.g. `"SBS96"`, `"ID83"`,
`"DBS78"`).

```{r}
# BASCULE
sigs <- get_tumourevo_signatures("SPN04", 50, 0.6, "mutect2", "sequenza",
                                  "BASCULE")
sigs$refined_fit
sigs$base_fit

# SigProfiler
sigs <- get_tumourevo_signatures("SPN04", 50, 0.6, "mutect2", "sequenza",
                                  "sigprofiler", context = "SBS96")
sigs$context_matrix
sigs$COSMIC_exposure
sigs$denovo_signatures
```
